---
title: "morrow_plots_v02"
output: html_document
date: "2024-08-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = "**")
library(tidyverse)
library(lubridate)
```

## Import v01 data

Source: Morrow Plots Data Curation Working Group (2022): Morrow Plots Treatment and Yield Data. University of Illinois at Urbana-Champaign. https://doi.org/10.13012/B2IDB-7865141_V1

```{r read}
mp01 <- read_csv("morrow-plots_v01_1888-2021_soil.csv")

```
## Create v02 with corrected 2021 all_corn data

The T/F all_corn variable flags years when corn was grown on all three plots. 2021 was an all-corn year, but the data for that year was incorrectly flagged as FALSE.

```{r allcorn}
## create a new version of the data with corrected values for 2021 all_corn
mp02 <- mp01 %>% 
  mutate(all_corn = 
    (if_else(year==2021,
             TRUE,
             all_corn)
    ))
```

v02 data will be made available in the Illinois Data Bank at https://doi.org/10.13012/B2IDB-7865141_V2