---
title: "Visualizing Morrow Plot Data (1888-2021)"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = "**")
library(tidyverse)
library(lubridate)
## adding cowplot for more viz options including custom grids
library(cowplot)
library(scales)


```


### Setup
Read in Morrow 1888-2021 v02

```{r read}
mp882103 <- read_csv("Morrow_1888-2021_v02.csv")
```


Set plot_dir as a factor so we can order it according to direction instead of alphabetically.

```{r as factor}
mp882103 <- mp882103 %>% 
  mutate(plot_dir = factor(plot_dir, levels = c("NW", "NE", "SW", "SE")))
```


Add phase column using phases identified in Wander that correspond to changes in treatment strategies.

```{r phase}
mp882103 <- mp882103 %>% 
  mutate(phase = case_when(year <= 1903 ~ 1,
                           year >= 1904 & year <= 1954 ~ 2,
                           year >= 1955 & year <= 1967 ~ 3,
                           year >= 1968 & year <= 1997 ~ 4,
                           year >= 1998~ 5)
           ) %>% 
  select(phase, everything())
```


Subset corn crops

```{r sub corn}
## subset corn crops
corn <- filter(mp882103, corn == TRUE) %>% 
  mutate(rotation = as_factor(rotation))
```


Set up reusable functions for common chart elements.

```{r elements}


## x axis breaks by decade
decades <- scale_x_continuous(breaks= c(1888,1900,1920, 1940,1960,1980,2000,2020))

## phase lines
phases <- tibble(phase = 1:5, date = c(1888,1904,1955,1968,1998), y = .66)
lines <- geom_vline(data = phases, mapping = aes(xintercept=date), color = "dimgray", size = 1)

## phase labels
labels <- geom_label(data = phases, mapping = aes(x = date, y = y, label = date), color = "dimgray")
                            
## facet by direction
direction <-  facet_wrap(vars(plot_dir), ncol = 2)

## facet by subplot
subplot <-  facet_wrap(vars(plot), ncol = 4)

## crop names on y axis
cropnames <- scale_y_discrete(labels=c("Alfalfa", "Corn in Rotation", "Continuous Corn", "Hay", "Oats", "Soybeans"))

## crop color code
cropcolor <-   scale_color_manual(
  labels = c("A" = "Alfalfa",
                                "C" = "Corn",
                                "CC"="Cont. Corn",
                                "H" = "Hay",
                                "O" = "Oats",
                                "S" = "Soybeans"),
  values = c("A" = "#88CCEE",
                                "C" = "#44AA99",
                                "CC"="#999933",
                                "H" = "#DDCC77",
                                "O" = "#cc6677",
                                "S" = "#AA4499"))

## treatment color code
treatcolor <-   scale_color_manual(values = c(
                                "none" = "#88CCEE",
                                "MLrP" = "#332288",
                                "MLbP" = "#44AA99",
                                "MLrPNsPK" = "#cc6677",
                                "0LNsPK" = "#DDCC77",
                                "MLP" = "#117733",
                                "LHNPK" = "#AA4499",
                                "LNPK" = "#999933"
                                ))

## treatment shape
treatshape <-   scale_shape_manual(values = c(
                                "FALSE" = 1,
                                "TRUE" = 16))


## treatment linetype
treatline <- scale_linetype_manual(values = c("dashed", "solid"))


```


## Crop History

```{r crop time}

ggplot() +
  geom_count(data = mp882103, mapping = aes(x= year, y= crop, color = crop, size = 1))+
  decades+
  lines+
  labels+
  cropcolor+
  cropnames+
  labs(title = "Crop Timeline",
       subtitle = "Years each crop was grown") +
  theme_light()+
  theme(legend.position = "none")+
   guides(x = guide_axis(angle = 45)) 


```

```{r crop facet}

ggplot() +
  geom_count(data = mp882103,
             mapping = aes(x= year, y= crop, color = crop, size = 1))+
  decades+
  lines+
  cropcolor+
  cropnames+
    labs(title = "Crop Timeline by Plot Number",
       subtitle = "Years each crop was grown",
       caption = "Plot 3 = continuous corn   |   Plot 4 = two-crop rotation   |   Plot 5 = three-crop rotation") +
  theme_light()+
  theme(legend.position = "none")+
   guides(x = guide_axis(angle = 45))+
  facet_grid(plot_num ~ .)


```


## Treatment History

```{r treat timeline}

## setting treatments as a factor and ordering the factor values according to phase so they sort chronologically.
mp882103 <- mutate(mp882103, treatment = fct_reorder(treatment, desc(phase)))


ggplot() +
  geom_count(data = mp882103, mapping = aes(x= year, y= treatment, color = treatment, size = 1), shape = 15)+
  decades+
  lines+
  labels+
  treatcolor+
    labs(title = "Treatment Timeline",
       subtitle = "Years each treatment was applied") +
  theme_light()+
  theme(legend.position = "none")+
   guides(x = guide_axis(angle = 45)) 

```


## Treatment Maps

Faceting revealed that phase 4 treatments began in 1968 not 1967 as indicated in Wander. Changed the phase variable to start phase 4 in 1968, but should be investigated.

```{r treat facet}


treatphase <- function(z) {
  
ggplot(data = filter(mp882103, phase == z & plot_num == 4), 
       mapping = aes(x= year, y= treatment)) +
  geom_line(aes(color = treatment, size = 1))+
  geom_label(aes(label = treatment),
            data = filter(mp882103, phase == z & plot_num == 4) %>% filter(year == min(year)),
            nudge_x = 5)+
  scale_x_continuous(labels = label_number(accuracy = 1))+
  labs(title = paste("Phase", z, "Treatment Map"),
       subtitle = "Years each treatment was applied",
       caption = "N & S indicate North and South. Western sections are split into A & B. Eastern sections are split into C & D.")+
  theme_light()+
  theme(legend.position = "none", axis.text.y=element_blank())+
   guides(x = guide_axis(angle = 45))+
  
  treatcolor+
  subplot
}


treatphase(1)
treatphase(2)
treatphase(3)
treatphase(4)
treatphase(5)
```


## Yield History by Plot

Scatterplot for each plot with yield_bush data color coded by crop. Excludes hay, which is measured in tons.

```{r yield history}

plotyield <- function(n) {

ggplot(data = filter(mp882103, !is.na(crop) & plot_num == n)) +
  geom_point (mapping = aes(x= year, y = yield_bush, color = crop, shape = treated), size = 2.5, na.rm = TRUE) +
  treatshape+
  cropcolor+
  lines+
  theme_light()+
  guides(x = guide_axis(angle = 45)) +
             labs(title = paste("Yield History Plot", n)) +
  facet_wrap(vars(plot_dir), ncol = 2) 
}

plotyield(3)
plotyield(4)
plotyield(5)


```


## Corn Yield Comparison

```{r corn}

ggplot(data = corn, na.rm = TRUE, aes(year, yield_bush, color = rotation)) +
  geom_smooth(aes(linetype = treated))+
  guides(x = guide_axis(angle = 45)) +
  treatline+
  decades+
  lines +
  labels+
  labs(title = "Corn Yield by Rotation and Treatment",
       subtitle = "The highest yields consistently come from corn that is both rotated and treated.")+
  theme_light()
  
  

```


### Create summary dataset of corn yield by rotation, treatment, phase

```{r summary}
summary <- group_by(corn, rotation, treated, phase) %>% 
  summarize(yield_min = min(yield_bush, na.rm = TRUE),
            yield_ave = mean(yield_bush, na.rm = TRUE),
            yield_max = max(yield_bush, na.rm = TRUE)) %>% 
  arrange(rotation)

write_csv(summary, "Morrow_1888-2021_v03_summary.csv")
```


### Export merged and transformed data 1888-2021_v03

```{r export}
write_csv(mp882103, "Morrow_1888-2021_v03.csv")
```













